\added{1}{Permutation groups, which are groups whose elements are realised as permutations of some (nonempty) set $\Omega$, play a central role in \textbf{computational group theory}, which uses computational methods to compute and store information about groups. The notion of a \textit{group action} arises naturally from the concept of permutation groups, since group actions encode group elements as permutations: every permutation group on $\Omega$ naturally induces an action on $\Omega$, and a permutation group arises from every group action. Let $G$ be a group and $\alpha \in \Omega$; the notions of an \textit{orbit} and \textit{stabiliser} of $\alpha$ can be defined for $G$-actions on $\Omega$. \added{2}{(The stabiliser of $\alpha$ comprises elements $g \in G$ whose associated permutations fix $\alpha$.)} A \textit{base} is an ordered list $B$ of elements of $\Omega$ such that only the identity element $1 \in G$ fixes every element of $B$ pointwise; a base gives rise naturally to a series of stabiliser subgroups called the \textit{stabiliser chain}, and a generating set that respects this structure is called a \textit{strong generating set}; these are used in numerous algorithms across computational group theory. We now approach this formally.}

\added{1}{
    \begin{definition}\label{def:permutation_group}
        Let $\Omega$ be a set. The \textbf{symmetric group} $\Sym(\Omega)$ on $\Omega$ consists of \textit{permutations} of $\Omega$ (bijections $\Omega \to \Omega$) under composition. A \textbf{permutation group} on $\Omega$ is a subgroup of $\Sym(\Omega)$; its \textbf{degree} is $|\Omega|$.
    \end{definition}}

\added{1}{When $\Omega = [n] := \{1,\dotsc,n\}$, we instead write $\Sym(n)$ for the \textit{symmetric group} of degree $n$.} \added{3}{(A similar notation is $\Alt(n)$ for the \textit{alternating group} of degree $n$ of even permutations of $n$.)} In this \thesis{}, we use the following conventions: for $\alpha \in \Omega$ and $g,h \in \Sym(\Omega)$, write $\alpha^g$ for the image of $\alpha$ under $g$ (instead of $g(\alpha)$ or $g \alpha$), and we perform $g$ then $h$ in the composition (product) $gh$, so that $\alpha^{gh} = (\alpha^g)^h$. \added{1}{Also, for the trivial subgroup we write $1 = \{1\} \leq G$.}

As permutations are a central object of study, we recall \added{1}{the} \textit{disjoint cycle notation}: let $\Omega$ be a set. Then if $\alpha_1,\dotsc,\alpha_r$ are distinct elements of $\Omega$, we write $\sigma = (\alpha_1,\alpha_2,\dotsc,\alpha_r)$ for the cycle $\sigma \in \Sym(\Omega)$ with $\alpha_i^\sigma = \alpha_{i+1}$ for $1 \leq i < r$, $\alpha_r^\sigma = \alpha_1$, and $\alpha^\sigma = \alpha$ for $\alpha \in \Omega \setminus \{\alpha_1,\dotsc,\alpha_r\}$. If $\Omega$ is finite, then every permutation can be decomposed into a product of disjoint cycles; we generally omit fixed points (cycles of length 1). \added{1}{The lengths of these cycles give the \textit{cycle type} of $\sigma$. The identity permutation $\Id_\Omega$ is written as $1_{\Sym(\Omega)}$, $()$, or just $1$ when context is clear.}

\added{3}{A group $G$ is \textbf{generated by a set} $X$ if every $g \in G$ can be written as} $g = x_1^{\varepsilon_1}\dotsb x_k^{\varepsilon_k}$ with each $x_i \in X$, $\varepsilon_i = \pm 1$, and $k \in \N$; if $k = 0$ then we have the empty product $g = 1$. We write $G = \langle X \rangle$. When \added{2}{$G = \langle X \rangle$ for some finite $X$}, $G$ is \textbf{finitely generated}. \added{3}{When $X,Y \subseteq G$, we write $\langle X,Y \rangle = \langle X \cup Y \rangle \leq G$ as the subgroup generated by $X$ and $Y$; this notion naturally extends to any finite number of subsets of $G$.}

% A \textbf{group presentation} of a group $G$ is a nonempty set $X$ of \textbf{generators} and a set $R$ of \textbf{relations} such that $G = \langle X \rangle$ and $R$ is a subset of the free group $F_X$ on $X$ such that for any relator $r \in R$, $r = 1$ in $G$, and we can write $r = y_1^{\varepsilon_1}\dotsb y_\ell^{\varepsilon_\ell}$ with $\ell \in \N$ and each $y_i \in X$ and $\varepsilon_i = \pm 1$. We write $G = \langle X \mid R \rangle$; we can see that $G$ is isomorphic to a quotient of the free group $F_X$ by the smallest normal subgroup containing $R$.

\section{Group actions}

\input{group_concepts/group_actions.tex}

\section{Bases and stabiliser chains}\label{sec:bases_stabiliser_chains}

\input{group_concepts/bases_stabiliser_chains.tex}

\section{Overview of classification results}

\input{group_concepts/classification_results.tex}