Firstly, I would like to express my gratitude to my supervisors Heiko and Santiago, who provided me with guidance and direction. I appreciate their tireless effort in imparting their knowledge, and for the hours spent reading over my \thesis{} to give prompt and useful feedback, even on weekends and late at night. I would also like to thank Norm, Ian, Nick, Wenhui, Andrea, Todd, Jessica, Tim and Eric, whose courses I have learned and received much from, and for the valuable feedback from coursework that has shaped my \thesis{}.

I would also like to thank my friends in the Honours cohort, especially Jeremy, Will and Clayton, for their friendship and support over the year, for all the good memories, and for giving me useful tips for my \thesis{}. For Puti too, who has helped me so much. This extends to various members of online communities -- the Maths @ Monash and Chats \& Reacts Discord servers, whose company I have been blessed with over this year.

I would like to thank my parents for the countless hours they have spent supporting me in every way at home to give me time and energy to maintain my busy schedule. I would also like to thank my friend Wes for being my greatest support through good and trying times this year, and giving me inspiration and assistance for my Honours presentation on the Rubik's group of permutations. Also, to all my church group members and fellow leaders for their constant support and care in busy times.

Lastly, I would like to thank God for His guiding hand in my life, and for being the reason I am where I am today. In the midst of busyness and challenges this year, His presence has given me hope, rest and security. I am so grateful for His constant love, faithfulness, mercy and grace, and I thank Him for this Honours experience in this season of my life:

\begin{quote}
    ``There is a time for everything, \\
    \hspace*{15pt} and a season for every activity under the heavens'' \\
    \null\hfill --- Ecclesiastes 3:1 (NIV)
\end{quote}